---
title: "icdmerge: Quickstart (Toy Example)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{icdmerge: Quickstart (Toy Example)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4
)
```
## What this package does

icdmerge merges fine-grained ICD diagnosis nodes into clinically interpretable merged nodes under a clinical constraint:

- Each disease is represented by a global structural profile (co-occurrence profile against all diseases).

- Clustering and merging are done within each block only (clinical interpretability).

- The package selects a block-specific number of clusters K* using consensus k-means and PAC.

This vignette demonstrates the workflow using a toy dataset with explicit comorbidity structure so that merging is likely to happen.

## Generate a toy dataset with comorbidity structure

The toy generator below creates:

- 3 blocks (Block_A, Block_B, Block_C)

- Each block has 8 diseases (total 24 diseases)

- Within each block, diseases form two latent subgroups with stronger within-group comorbidity

- A global comorbidity factor introduces network-wide co-occurrence patterns (global roles)

```{r toy_generator}
library(data.table)

simulate_icdmerge_toy <- function(
  n_patients = 800,
  blocks = c("Block_A", "Block_B", "Block_C"),
  diseases_per_block = 8,
  seed = 1
) {
  set.seed(seed)

  # disease IDs and block mapping
  disease_ids <- unlist(lapply(seq_along(blocks), function(i) {
    sprintf("D%s_%02d", LETTERS[i], 1:diseases_per_block)
  }))
  block_vec <- rep(blocks, each = diseases_per_block)

  # latent factors per patient
  g  <- rnorm(n_patients)  # global comorbidity tendency
  bA <- rnorm(n_patients)  # block-level tendencies
  bB <- rnorm(n_patients)
  bC <- rnorm(n_patients)

  # subgroup tendencies within each block (2 subgroups)
  sA1 <- rnorm(n_patients); sA2 <- rnorm(n_patients)
  sB1 <- rnorm(n_patients); sB2 <- rnorm(n_patients)
  sC1 <- rnorm(n_patients); sC2 <- rnorm(n_patients)

  # helper to get block factor by name
  block_factor <- function(blk) {
    if (blk == "Block_A") return(bA)
    if (blk == "Block_B") return(bB)
    bC
  }

  # baseline prevalence around ~5%-25% depending on disease
  base <- runif(length(disease_ids), min = -2.8, max = -1.2)

  # weights: tune these to make co-morbidity/merging likely
  w_global <- 0.9
  w_block  <- 1.2
  w_sub    <- 2.0  # ↑ slightly stronger within-subgroup comorbidity to increase merge probability

  # assign subgroup within each block: first half subgroup1, second half subgroup2
  stopifnot(diseases_per_block %% 2 == 0)
  subgroup <- rep(rep(c(1, 2), each = diseases_per_block / 2), times = length(blocks))

  # mild cross-block "role" signal
  role_shift <- rep(0, length(disease_ids))
  role_shift[block_vec == "Block_A" & subgroup == 1] <- 0.20
  role_shift[block_vec == "Block_B" & subgroup == 1] <- 0.10
  role_shift[block_vec == "Block_C" & subgroup == 2] <- 0.15

  # generate diagnoses then convert to long form
  eid <- sprintf("P%04d", 1:n_patients)
  out_list <- vector("list", length(disease_ids))

  for (j in seq_along(disease_ids)) {
    blk <- block_vec[j]
    bf  <- block_factor(blk)

    # subgroup factor by block and subgroup
    sf <- switch(
      blk,
      Block_A = if (subgroup[j] == 1) sA1 else sA2,
      Block_B = if (subgroup[j] == 1) sB1 else sB2,
      Block_C = if (subgroup[j] == 1) sC1 else sC2
    )

    # linear predictor -> probability via logistic
    eta <- base[j] + w_global * g + w_block * bf + w_sub * sf + role_shift[j]
    p <- 1 / (1 + exp(-eta))

    y <- rbinom(n_patients, size = 1, prob = p)

    if (sum(y) > 0) {
      out_list[[j]] <- data.table(
        eid = eid[y == 1],
        Category_No = disease_ids[j]
      )
    } else {
      out_list[[j]] <- NULL
    }
  }

  raw_data <- rbindlist(out_list, use.names = TRUE, fill = TRUE)
  raw_data <- unique(raw_data)

  codebook <- data.table(
    No = disease_ids,
    block = block_vec,
    Codes = disease_ids,
    Names = paste("Toy disease", disease_ids)
  )

  # Frequency = number of unique patients with the disease
  freq_dt <- unique(raw_data)[, .(Frequency = uniqueN(eid)), by = Category_No]
  codebook <- merge(codebook, freq_dt, by.x = "No", by.y = "Category_No", all.x = TRUE)
  codebook[is.na(Frequency), Frequency := 0L]

  list(raw_data = raw_data, codebook = codebook)
}

toy <- simulate_icdmerge_toy()

# Avoid printing huge tables in the vignette
head(toy$raw_data)
head(toy$codebook)
```

A quick look at disease frequencies:
```{r}
toy$codebook[, summary(Frequency)]
toy$codebook[, .(n_diseases = .N, minF = min(Frequency), maxF = max(Frequency)), by = block]

```

## Run icdmerge on the toy data

We keep parameters small so the vignette runs fast, but large enough to demonstrate merges.

```{r run_icdmerge}
library(icdmerge)

res <- run_block_merge(
  raw_data = toy$raw_data,
  codebook  = toy$codebook,
  prevalence_cutoff = 0.01,   # toy data: keep common diseases
  n_resamples = 60,           # keep moderate; should finish quickly
  k_max_per_block = 6,
  write_files = FALSE,
  verbose = TRUE
)

print(res)
```

## Inspect selected K* and merge results
```{r}
#Block-wise K* distribution
kstar <- res$kstar
kstar[, .N, by = K_star][order(K_star)]
kstar[order(-n_nodes)][1:10]

#How many merges happened?
res$lineage[Member_Count > 1, .N]
res$lineage[Member_Count > 1][order(-Member_Count)][1:10, .(Current_ID, Block, Member_Count)]

#Merged map
head(res$update_map)

```

## Notes and tips

- If your toy run produces no merges, increase comorbidity strength by raising w_sub in the generator (e.g., from 1.8 to 2.2), or increase n_resamples.
- For real data, the recommended workflow is:
  1. Ensure your raw_data and codebook meet the required column names
  2. Run run_block_merge() with a stable prevalence_cutoff
  3. Review PAC_Summary_Kstar.csv (or res$kstar) to understand block-wise granularity
  4. Use the merge dictionary (Final_Merged_Dict.csv) to preserve interpretability
